Pan-cancer analysis

### 2021-03-10 src/manuscript_figures.Rmd
collect figures into a list to print them later

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/settings.yaml')
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
theme_set(ggpubr::theme_pubclean())

inputDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/output'
assay_dir <- settings$assay_output$folder
surv <- do.call(rbind, readRDS(settings$surv))
projectDict <- sapply(yaml::read_yaml(settings$projectDict), 
                      function(x) gsub(' ', '', unlist(strsplit(x, ','))))
tcga_subtypes <- process_tcga_subtypes()
```

@@

Fig1 <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B')), 
                  plot_grid(p3, p4, rel_widths = c(5,3), nrow = 1, labels = c('C', 'D')), 
                  nrow = 2,
                  rel_heights = c(2,3))

print(Fig1)
ggsave('pancancer_cancer_type_prediction.pdf', plot = Fig1, units = 'in', width = 12, height = 9)

@@

Fig2 <- plot_grid(plot_grid(p1, p2, nrow = 2), p3, ncol = 2, rel_widths = c(2, 3), 
          labels = 'AUTO')
print(Fig2)

ggsave('pancancer_subtype_prediction.pdf', plot = Fig2, units = 'in', width = 12, height = 9)

# train/test accuracy correlation 
#ggscatter(pancancer_subtype_stats, x = 'train_auc', y = 'test_auc', 
 #         color = 'method', facet.by = 'method', add = 'reg.line')
```

@@

#M <- LFs$pca_factors[[e]]
#ct <- surv[match(rownames(M), bcr_patient_barcode)]$project

@@

Fig <- cowplot::plot_grid(p1$p, p2$p, p3$p, p4, nrow = 4, labels = 'AUTO')
print(Fig)
ggsave("pancancer.unsupervised_clustering.pdf", Fig, units = 'in', width = 10, height = 12)
```

@@

Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs 
```{r clustering_metrics, fig.width=10, fig.height=8}
multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments'))
clusters <- sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
    require(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      Mr <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
      M <- t(subset_features_by_variance(t(Mr), topN = 10))
     } else {
      M <- M[,head(colnames(M), 10),drop = F]
    }
    val <- c('internal') 
    if(ncol(M) > 2) {val <- c('internal', 'stability')}
    res <- clValid::clValid(obj = M,
                          nClust = 3:8,
                          clMethods = 'kmeans',
                          validation = val,
                          maxitems = 10000,
                          nstart = 1000)
    memberships <- as.data.frame(do.call(cbind, lapply(res@clusterObjs[['kmeans']], function(x) x[['cluster']])))
    colnames(memberships) <- paste0('k', colnames(memberships))
    stats <- as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
    colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
    return(list('clustering_stats' = stats, 'cluster_memberships' = memberships))
  })
})
parallel::stopCluster(cl)

@@

Fig4 <- plot_grid(plot_grid(p1, p2, labels = 'AUTO', ncol = 1),
                  p3, ncol = 2, rel_widths = c(2,3), labels = c('','C'))
print(Fig4)
ggsave('cluster_stability_metrics.pdf', plot = Fig4, units = 'in', 
       width = 10, height = 8)

@@

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 'experiments'))
surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    require(data.table)
    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
    res <- get_prognostic_covariates(clin, M)
    if(!is.null(res)) {
      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
      res$project <- cancer_type
      v <- res[order(pval)][1]$variable
      dat <- merge(clin, M[,v,drop = F], by = 'row.names')
      rownames(dat) <- dat$Row.names
      dat$Row.names <- NULL
      # compute hazard ratios 
      cox1 <- coxph(Surv(PFI.time,PFI) ~ ., 
              data=dat, x = TRUE)
      hr <- data.frame(summary(cox1)[['conf.int']]) # get hazard ratios with conf intervals
      hr <- hr[grep(v, rownames(hr)), c(1,3,4)]
      colnames(hr) <- c('hr', 'hr_low', 'hr_high')
      rownames(hr) <- v
      p1 <- survminer::ggforest(cox1, data = dat)

@@

plots <- sapply(simplify = F, names(surv.stats), function(x) {   
  p <- surv.stats[[x]][['p2']] 
  if(!is.null(p)) { 
    p + theme(axis.text.x = element_text(size = 8), 
              axis.text.y = element_text(size = 8), 
              axis.title.x = element_blank(), 
              axis.title.y = element_blank(), 
              legend.title = element_blank(),
              legend.position = c(0.5, 0.9),
              legend.text = element_text(size = 8), 
              legend.background = element_blank(), 
              legend.direction = 'vertical') 
    }
})
plots <- plots[!sapply(plots, is.null)]
p1 <- cowplot::plot_grid(plotlist = plots, nrow = 3, 
                   labels = sapply(strsplit(names(plots), '\\.'), function(x) x[1]), 
                   label_size = 8)
print(p1)
ggsave(filename = 'surv.adjustedcurves.prog_lfs.pdf', plot = p1, width = 12, height = 8, units = 'in')

# plot hazard ratios for top LF per cancer type 
hr <- do.call(rbind, lapply(names(surv.stats), function(x) {
  if(is.null(surv.stats[[x]])) {return(NULL)}
  message(x)
  df <- surv.stats[[x]][['hr']]
  df$project <- unlist(strsplit(x, '\\.'))[[1]]
  return(df)
}))
hr$latent_factor <- rownames(hr)
# remove hr with Inf values
hr <- hr[!is.infinite(hr$hr_high),]

p2 <- ggplot(hr, aes(x = reorder(paste0(project, ' (', latent_factor, ')'), hr), y = log(hr))) +
  geom_bar(stat = 'identity') + 
  geom_errorbar(aes(ymin = log(hr_low), ymax = log(hr_high)), width = 0.2) + 
  #geom_text(aes(label = latent_factor, y = ifelse(hr > 1, log(hr+0.005), log(hr-0.005))), 
   #         nudge_x = 0.25) + 
  labs(x = '', y = 'log(hazard-ratio) of top latent factor per cancer\n(adjusted for tumor stage + age + gender)')+
  coord_flip()

@@

p <- cowplot::plot_grid(p3, p2, nrow = 1, labels = 'AUTO')

print(p)
ggsave(filename = 'surv.Cindex_improv.hazard_ratio.pdf', plot = p, width = 12, height = 8, units = 'in')

@@

```{r clinical_factors, fig.width=12, fig.height=9}
experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
# find clinical-factor associated LFs 
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('experiments', 'LFs'))
clin.stats <- do.call(rbind, 
        pbapply::pbsapply(cl = cl, simplify = F, 
               experiments, function(experiment) {
               require(data.table)
               source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
               M <- LFs$maui_factors[[experiment]]
               clin <- process_clinical_covariates(rownames(M))
               dt <- do.call(rbind, 
                    pbapply::pbsapply(simplify = F, 
                                      colnames(clin), function(x) {
                      if(!is.numeric(clin[,x])) {
                        factors <- clin[,x]
                        m <- M[!is.na(factors),]
                        factors <- factors[!is.na(factors)]
                        dt <- get_factor_specific_variables(m, factors) #compute_anova(m, factors)
                        dt$label <- x
                        return(dt)
                      }
                      return(NULL)
                      }))
               dt$project <- unlist(strsplit(experiment, "\\."))[[1]]
               return(dt)
             }))
parallel::stopCluster(cl)

@@

# pick example cases and show more detailed LF activation 
# top examples by clinical factor type
# top 5 variables per clinical factor

top <- clin.stats[padj < 0.05,.SD[which.min(padj)], by = label][order(padj),.SD[which.min(padj)],by = project]
plots <- sapply(simplify = FALSE, 1:4, function(i) {
  pr <- top$project[i]
  l <- top$label[i]
  M <- LFs$maui_factors[[paste0(pr, '.cnv_gex_meth_mut')]]
  factors <- process_clinical_covariates(rownames(M))[[l]]
  M <- M[!is.na(factors),]
  factors <- factors[!is.na(factors)]
  plot_label_specific_vars(M, factors, string_width = 24)+ #, top_vars = top_vars) + 
    labs(title = pr, subtitle = l) + 
    theme(legend.position = 'none')
})

p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))
p <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
print(p)
ggsave(filename = 'clinical_factors.LFs.pdf', p, units = 'in', width = 12, height = 9)
```

# Predicting MSI status using latent factors
```{r msi_stats, fig.width=10, fig.height=8}
# predict MSI status for pan-gastrointestinal tumors
# read MSI data
dataDir <- '/data/local/buyar/pancancer_diagnostics/data/GDCdata_prepared'
MSI <- do.call(rbind, lapply(projectDict$`pan-gastrointestinal`, function(pr) {
  f <- file.path(dataDir, paste0(pr, '.MSI.tsv'))
  dt <- data.table::fread(f)[,c(2,4)]
  colnames(dt)[2] <- 'MSI'
  dt$project <- pr
  return(dt)
}))
# remove samples with undertermined MSI status
MSI <- MSI[MSI %in% c('MSI-H', 'MSI-L', 'MSS')]
msi_stats <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][['pan-gastrointestinal.cnv_gex_meth_mut']]
  y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]
  #res <- predict_cluster_labels(M, y, partition = 0.6, num.trees = 2000)
  res <- predict_cluster_labels_glmnet(M, y)
  #res <- predict_labels_glmnet(M, y)
  res$method <- tool
  return(res)
}))
msi_stats$method <- toupper(gsub('_factors', '', msi_stats$method))
# with glmnet and error bars 
# for supplements (includes other tools)
# p1 <- ggplot(msi_stats, aes(x = method, y = auc)) + 
#   geom_bar(aes(fill = method), stat = 'identity') + 
#   geom_errorbar(aes(ymin = auc.lower, ymax = auc.upper), width = 0.2) + 
#   labs(y = 'AUC',x = '') + theme(legend.title = element_blank()) + 
#   facet_grid(~ label)

@@

p2.1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(group))) + 
  theme(legend.title = element_blank())
p2.2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
  geom_point(aes_string(color = top_lf)) + 
  scale_color_gradient(low = 'gray', high = 'red')
p2.3 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(cancertype))) + 
  theme(legend.title = element_blank())
p2.4 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(subtype))) + 
  theme(legend.title = element_blank())

print(cowplot::plot_grid(p2.1, p2.2, p2.3, p2.4, ncol = 2, labels = 'AUTO'))
# top features that contribute to top factors relevant to MSI status prediction
f <- data.table::fread(file.path(inputDir, 'maui_factors', 
                               paste('pan-gastrointestinal', # project 
                                      'cnv_gex_meth_mut', #data type
                                      'maui_factors', # tool
                                      'feature_weights.csv', sep = '.')))
feat.M <- as.matrix(f[,-1])
rownames(feat.M) <- gsub('assay: ', '', f$V1)

@@

p <- cowplot::plot_grid(p1, p3, p2.1, p2.2, labels = 'AUTO')
print(p)
ggsave(filename = 'MSI.status.pdf', p, units = 'in', width = 10, height = 8)

@@

### Use-case: pan-gastrointestinal cancers
```{r semi_supervised_subtypes, fig.width=10, fig.height=12}
require(survminer)
require(cluster)
tcga_subtypes <- process_tcga_subtypes()
# subtype specific LFs
get_top_by_label <- function(dt, label.type, topN = 5) {
  do.call(rbind, lapply(split(dt, dt[,get(label.type)]), function(x) {head(x, topN)}))
}

pr <- 'pan-gastrointestinal' #'TCGA-BRCA' # 'glioma' 
experiment <- paste0(pr, '.cnv_gex_meth_mut')
M <- LFs$maui_factors[[experiment]]
st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
# find subtype specific LFs 
dt <- get_factor_specific_variables(M, st)

# use gap statistic to get clusters
vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 3)[['variable']])
# use gap statistic to find optimum clusters 
#r1 <- cluster::clusGap(M[,vars], FUNcluster = kmeans, K.max = 20, iter.max = 50, B = 100)
#plot(r1)
#k <- cluster::maxSE(r1$Tab[,3], r1$Tab[,4], method = 'firstSEmax', SE.factor = 1)
k <- length(na.omit(unique(st)))
# cluster samples using only subtype specific factors 
clust <- data.frame(get_kmeans_subtypes(M[,vars], k_vals = k, min_cluster_size = 0))

@@

# use hallmark gene sets to annotate cluster specific functions
scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_hallmarks/TCGA.ssgsea_all.RDS'))
S <- scores[intersect(rownames(M), rownames(scores)),]
clust$tcga <- tcga_subtypes[match(rownames(clust), bcr_patient_barcode)]$Subtype_Selected

@@

# altogether
P <-  cowplot::plot_grid(p1, p2, p3, p4, h1$gtable, h2$gtable, 
                         labels = 'AUTO', ncol = 2)
print(P)
ggsave(filename = paste0('semi_supervised.clustering.', experiment,'.pdf'), 
       P, units = 'in', 
       width = 10, height = 12)

@@

Pick survival-predictive LFs and cluster samples based on those
```{r semi_supervised_survival, fig.width=10, fig.height=12}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage', 
                  'PFI', 'PFI.time')

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 'experiments'))
prog.clust <- pbapply::pbsapply(cl = cl, simplify = F, experiments, 
                                function(experiment){
    require(data.table)
    require(survminer)
    source('/data/local/buyar/arcas/subtyping_paper/src/common_functions.R')
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
    res <- get_prognostic_covariates(clin, M)
    # now cluster samples based on only survival predictive factors
    prog_lfs <- unique(res$variable)
    if(length(prog_lfs) < 2) {
      return(NULL)
    }
    clust <- get_spectral_clusters(M[,prog_lfs], k_vals = 2:10)
    # find value of k which minimizes log-rank p-value
    dt <- do.call(rbind, lapply(colnames(clust), function(x) {
      message(x)
      pval <- plot_survival(surv, clust[,x,drop=F], return_pval = T)
      data.table('k' = x, 'pval' = pval)
    }))

@@

# use hallmark gene sets to annotate cluster specific functions
#scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_hallmarks/TCGA.ssgsea_all.RDS'))

# use canonical pathways to annotate prognostic groups
scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_canonical_pathways/ssgsea.canonical_pathways.RDS'))

# scores geneset scores
# M survival predictive LF matrix
# maui_clust: best survival prognostic clustering of M 
plot_clusts <- function(M, maui_clust, scores) {
  df <- plot_tsne(M, as.factor(maui_clust[,1]), returnData = T)
  colnames(df)[3] <- 'maui'
  #df$tcga <- tcga_subtypes[match(rownames(df), bcr_patient_barcode)]$Subtype_Selected

@@

Pc <- cowplot::plot_grid(p1, plot_grid(P$p1, P$p2, nrow = 1, labels = c('B', 'C')), 
                   P$h1$gtable, nrow = 3, labels = c('A', '', 'D'))

print(Pc)
ggsave(filename = paste0('semi_supervised.prognostic.clustering.', experiment,'.pdf'), 
       Pc, units = 'in', 
       width = 10, height = 12)

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# todo: pass this as argument
settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/settings.yaml')
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
theme_set(ggpubr::theme_pubclean())

# todo: move this to settings file
inputDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/output'
assay_dir <- settings$assay_output$folder
surv <- do.call(rbind, readRDS(settings$surv))
projectDict <- sapply(yaml::read_yaml(settings$projectDict), 
                      function(x) gsub(' ', '', unlist(strsplit(x, ','))))
tcga_subtypes <- process_tcga_subtypes()

Figures <- c() # accummulate all figures in this
SuppFigures <- c() # accumulate all supplementary figures in this

@@

Fig <- plot_grid(plot_grid(p1, p2, nrow = 1, labels = c('A', 'B')), 
                  plot_grid(p3, p4, rel_widths = c(5,3), nrow = 1, labels = c('C', 'D')), 
                  nrow = 2,
                  rel_heights = c(2,3))

Figures <- c(Figures, Fig)
ggsave('pancancer_cancer_type_prediction.pdf', plot = Fig, units = 'in', width = 12, height = 9)

@@

Fig <- plot_grid(plot_grid(p1, p2, nrow = 2), p3, ncol = 2, rel_widths = c(2, 3), 
          labels = 'AUTO')
print(Fig)

ggsave('pancancer_subtype_prediction.pdf', plot = Fig, units = 'in', width = 12, height = 9)

SuppFigures <- c(SuppFigures, Fig)
```

@@

@@

Fig <- cowplot::plot_grid(p1$p, p2$p, p3$p, p4, nrow = 4, labels = 'AUTO')
ggsave("pancancer.unsupervised_clustering.pdf", Fig, units = 'in', width = 10, height = 12)

Figures <- c(Figures, Fig)
```

@@

Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs 
```{r clustering_metrics, fig.width=10, fig.height=8}
multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
clusters <- sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      Mr <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
      M <- t(subset_features_by_variance(t(Mr), topN = 10))
     } else {
      M <- M[,head(colnames(M), 10),drop = F]
    }
    val <- c('internal') 
    if(ncol(M) > 2) {val <- c('internal', 'stability')}
    res <- clValid::clValid(obj = M,
                          nClust = 3:8,
                          clMethods = 'kmeans',
                          validation = val,
                          maxitems = 10000,
                          nstart = 1000)
    memberships <- as.data.frame(do.call(cbind, lapply(res@clusterObjs[['kmeans']], function(x) x[['cluster']])))
    colnames(memberships) <- paste0('k', colnames(memberships))
    stats <- as.data.table(res@measures)[V1 %in% c('Silhouette', 'APN')]
    colnames(stats) <- c('metric', 'k', 'clustering_method', 'score')
    return(list('clustering_stats' = stats, 'cluster_memberships' = memberships))
  })
})
parallel::stopCluster(cl)

@@

Fig <- plot_grid(plot_grid(p1, p2, labels = 'AUTO', ncol = 1),
                  p3, ncol = 2, rel_widths = c(2,3), labels = c('','C'))
ggsave('cluster_stability_metrics.pdf', plot = Fig, units = 'in', 
       width = 10, height = 8)

Figures <- c(Figures, Fig)
```

@@

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 
                                        'surv', 'experiments', 'settings'))
surv.stats <- pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    require(data.table)
    source(settings$utility_script)
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    clin <- get_surv_df(surv, clin[,intersect(colnames(clin), clin_factors),drop = F])
    res <- get_prognostic_covariates(surv, clin, M)
    if(!is.null(res)) {
      cancer_type <- strsplit(experiment, "\\.")[[1]][1]
      res$project <- cancer_type
      v <- res[order(pval)][1]$variable
      dat <- merge(clin, M[,v,drop = F], by = 'row.names')
      rownames(dat) <- dat$Row.names
      dat$Row.names <- NULL
      # compute hazard ratios 
      cox1 <- coxph(Surv(PFI.time,PFI) ~ ., 
              data=dat, x = TRUE)
      hr <- data.frame(summary(cox1)[['conf.int']]) # get hazard ratios with conf intervals
      hr <- hr[grep(v, rownames(hr)), c(1,3,4)]
      colnames(hr) <- c('hr', 'hr_low', 'hr_high')
      rownames(hr) <- v
      p1 <- survminer::ggforest(cox1, data = dat)

@@

plots <- sapply(simplify = F, names(surv.stats), function(x) {   
  p <- surv.stats[[x]][['p2']] 
  if(!is.null(p)) { 
    p + theme(axis.text.x = element_text(size = 8), 
              axis.text.y = element_text(size = 8), 
              axis.title.x = element_blank(), 
              axis.title.y = element_blank(), 
              legend.title = element_blank(),
              legend.position = c(0.5, 0.9),
              legend.text = element_text(size = 8), 
              legend.background = element_blank(), 
              legend.direction = 'vertical') 
    }
})
plots <- plots[!sapply(plots, is.null)]
p1 <- cowplot::plot_grid(plotlist = plots, nrow = 3, 
                   labels = sapply(strsplit(names(plots), '\\.'), function(x) x[1]), 
                   label_size = 8)
# plot hazard ratios for top LF per cancer type 
hr <- do.call(rbind, lapply(names(surv.stats), function(x) {
  if(is.null(surv.stats[[x]])) {return(NULL)}
  message(x)
  df <- surv.stats[[x]][['hr']]
  df$project <- unlist(strsplit(x, '\\.'))[[1]]
  return(df)
}))
hr$latent_factor <- rownames(hr)
# remove hr with Inf values
hr <- hr[!is.infinite(hr$hr_high),]

p2 <- ggplot(hr, aes(x = reorder(paste0(project, ' (', latent_factor, ')'), hr), y = log(hr))) +
  geom_bar(stat = 'identity') + 
  geom_errorbar(aes(ymin = log(hr_low), ymax = log(hr_high)), width = 0.2) + 
  labs(x = '', y = 'log(hazard-ratio) of top latent factor per cancer\n(adjusted for tumor stage + age + gender)')+
  coord_flip()

@@

Fig <- cowplot::plot_grid(cowplot::plot_grid(p3, p2, nrow = 1, labels = c('A', 'B')),
                        p1, labels = c('', 'C'), nrow = 2)

ggsave(filename = 'surv.Cindex_improv.hazard_ratio.pdf', plot = p, width = 12, height = 14, units = 'in')

Figures <- c(Figures, Fig)
```

@@

```{r clinical_factors, fig.width=12, fig.height=9}
experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
# find clinical-factor associated LFs 
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('experiments', 'LFs', 'settings'))
clin.stats <- do.call(rbind, 
        pbapply::pbsapply(cl = cl, simplify = F, 
               experiments, function(experiment) {
               require(data.table)
               source(settings$utility_script)
               M <- LFs$maui_factors[[experiment]]
               clin <- process_clinical_covariates(rownames(M))
               dt <- do.call(rbind, 
                    pbapply::pbsapply(simplify = F, 
                                      colnames(clin), function(x) {
                      if(!is.numeric(clin[,x])) {
                        factors <- clin[,x]
                        m <- M[!is.na(factors),]
                        factors <- factors[!is.na(factors)]
                        dt <- get_factor_specific_variables(m, factors) 
                        dt$label <- x
                        return(dt)
                      }
                      return(NULL)
                      }))
               dt$project <- unlist(strsplit(experiment, "\\."))[[1]]
               return(dt)
             }))
parallel::stopCluster(cl)

@@

# pick example cases and show more detailed LF activation 
# top examples by clinical factor type
# top 5 variables per clinical factor 
top <- clin.stats[padj < 0.05,.SD[which.min(padj)], by = label][order(padj),.SD[which.min(padj)],by = project]
plots <- sapply(simplify = FALSE, 1:4, function(i) {
  pr <- top$project[i]
  l <- top$label[i]
  M <- LFs$maui_factors[[paste0(pr, '.cnv_gex_meth_mut')]]
  factors <- process_clinical_covariates(rownames(M))[[l]]
  M <- M[!is.na(factors),]
  factors <- factors[!is.na(factors)]
  plot_label_specific_vars(M, factors, string_width = 24)+ 
    labs(title = pr, subtitle = l) + 
    theme(legend.position = 'none')
})

p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))
Fig <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
ggsave(filename = 'clinical_factors.LFs.pdf', p, units = 'in', width = 12, height = 9)
Figures <- c(Figures, Fig)
```

# Predicting MSI status using latent factors
```{r msi_stats, fig.width=10, fig.height=8}
# predict MSI status for pan-gastrointestinal tumors
# read MSI data
dataDir <- settings$GDCdataDir
MSI <- do.call(rbind, lapply(projectDict$`pan-gastrointestinal`, function(pr) {
  f <- file.path(dataDir, paste0(pr, '.MSI.tsv'))
  dt <- data.table::fread(f)[,c(2,4)]
  colnames(dt)[2] <- 'MSI'
  dt$project <- pr
  return(dt)
}))
# remove samples with undertermined MSI status
MSI <- MSI[MSI %in% c('MSI-H', 'MSI-L', 'MSS')]
msi_stats <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][['pan-gastrointestinal.cnv_gex_meth_mut']]
  y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]
  #res <- predict_cluster_labels(M, y, partition = 0.6, num.trees = 2000)
  res <- predict_cluster_labels_glmnet(M, y)
  #res <- predict_labels_glmnet(M, y)
  res$method <- tool
  return(res)
}))
msi_stats$method <- toupper(gsub('_factors', '', msi_stats$method))

@@

p2.1 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + geom_point(aes(color = as.factor(group))) + 
  theme(legend.title = element_blank())
p2.2 <- ggplot(df, aes(x = tSNE1, y = tSNE2)) + 
  geom_point(aes_string(color = top_lf)) + 
  scale_color_gradient(low = 'gray', high = 'red')

# top features that contribute to top factors relevant to MSI status prediction
f <- data.table::fread(file.path(inputDir, 'maui_factors', 
                               paste('pan-gastrointestinal', # project 
                                      'cnv_gex_meth_mut', #data type
                                      'maui_factors', # tool
                                      'feature_weights.csv', sep = '.')))
feat.M <- as.matrix(f[,-1])
rownames(feat.M) <- gsub('assay: ', '', f$V1)

@@

Fig <- cowplot::plot_grid(p1, p3, p2.1, p2.2, labels = 'AUTO')
ggsave(filename = 'MSI.status.pdf', Fig, units = 'in', width = 10, height = 8)

Figures <- c(Figures, Fig)
```

@@

### Use-case: pan-gastrointestinal cancers
```{r semi_supervised_subtypes, fig.width=10, fig.height=12}
require(survminer)
require(cluster)
pr <- 'pan-gastrointestinal' 
experiment <- paste0(pr, '.cnv_gex_meth_mut')
M <- LFs$maui_factors[[experiment]]
st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
# find subtype specific LFs 
dt <- get_factor_specific_variables(M, st)

vars <- unique(get_top_by_label(dt[padj < 0.05][order(padj)], 'ref_cl', 3)[['variable']])

k <- length(na.omit(unique(st))) # set k to number of known subtypes
# cluster samples using only subtype specific factors 
clust <- data.frame(get_kmeans_subtypes(M[,vars], k_vals = k, min_cluster_size = 0))

@@

# use hallmark gene sets to annotate cluster specific functions
# TODO: update this to point to /data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore
# and remove paths from script (put it in settings.yaml)
scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_hallmarks/TCGA.ssgsea_all.RDS'))
S <- scores[intersect(rownames(M), rownames(scores)),]
clust$tcga <- tcga_subtypes[match(rownames(clust), bcr_patient_barcode)]$Subtype_Selected

@@

# altogether
Fig <-  cowplot::plot_grid(p1, p2, p3, p4, h1$gtable, h2$gtable, 
                         labels = 'AUTO', ncol = 2)
ggsave(filename = paste0('semi_supervised.clustering.', experiment,'.pdf'), 
       Fig, units = 'in', 
       width = 10, height = 12)

Figures <- c(Figures, Fig)
```

@@

Pick survival-predictive LFs and cluster samples based on those
```{r semi_supervised_survival, fig.width=10, fig.height=12}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'clin_factors', 'surv', 'experiments'))
prog.clust <- pbapply::pbsapply(cl = cl, simplify = F, experiments, 
                                function(experiment){
    require(data.table)
    require(survminer)
    source(settings$utility_script)
    message(experiment)
    M <- LFs$maui_factors[[experiment]]
    # get clinical factors for the patients in matrix
    clin <- process_clinical_covariates(rownames(M))
    res <- get_prognostic_covariates(surv, clin, M, 'PFI', 'PFI.time')
    # now cluster samples based on only survival predictive factors
    prog_lfs <- unique(res$variable)
    if(length(prog_lfs) < 2) {
      return(NULL)
    }
    clust <- get_spectral_clusters(M[,prog_lfs], k_vals = 2:10)
    # find value of k which minimizes log-rank p-value
    dt <- do.call(rbind, lapply(colnames(clust), function(x) {
      message(x)
      pval <- plot_survival(surv, clust[,x,drop=F], return_pval = T)
      data.table('k' = x, 'pval' = pval)
    }))

@@

# use canonical pathways to annotate prognostic groups
# TODO: update this to point to /data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore
# and don't use hard-coded paths 
scores <- t(readRDS('/data/local/buyar/pancancer_diagnostics/data/geneset_scores/msigdb_canonical_pathways/ssgsea.canonical_pathways.RDS'))

# scores: geneset scores
# M: survival predictive LF matrix
# maui_clust: best survival prognostic clustering of M 
plot_clusts <- function(M, maui_clust, scores) {
  df <- plot_tsne(M, as.factor(maui_clust[,1]), returnData = T)
  colnames(df)[3] <- 'maui'

@@

Fig <- cowplot::plot_grid(p1, plot_grid(P$p1, P$p2, nrow = 1, labels = c('B', 'C')), 
                   P$h1$gtable, nrow = 3, labels = c('A', '', 'D'))

ggsave(filename = paste0('semi_supervised.prognostic.clustering.', experiment,'.pdf'), 
       Fig, units = 'in', 
       width = 10, height = 12)

Figures <- c(Figures, Fig)
